{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-03-28 paper.tex
Edit design considerations

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming language and system.

We contribute Denicek, a computational substrate that makes it easier to
implement the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. It provides three
core operations on edit histories: (i) edit application, (ii) merging of histories and (iii) conflict resolution.
We show that these operations serve as fundamental building blocks for a direct implementation
of a range of programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of an interactive data science notebook.
The case study shows that the Denicek computational substrate provides a robust basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

@@

A computational substrate defines the structures from which programs are constructed, how
the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

@@

What would be the ideal programming substrate for rich, interactive end-user programming systems?
In this paper, we aim to design a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup} including:

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Stage Methodology}
Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
a simple web-based end-user programming environment based on the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for testing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek (\S\ref{sec:case}), an end-user data
exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

\subsubsection*{Substrate}
The Denicek substrate brings together two central design ideas. First, it represents programs as
document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
static content. Second, Denicek does not store the document tree itself, but instead, it maintains
a sequence of edit operations through which the tree was constructed and transformed.

The substrate then provides three primitive operations for working with sequences of edits.
First, it can apply a series of edits to reconstruct a document. Second, it can merge two
diverging edit histories. Finally, it can detect conflicts when merging histories and, for
example, remove conflicting edits from one branch.

The key insight that we present in this paper is that many of the programming experiences can
be obtained by leveraging the uniform document representation alongside with a suitable combination
of the three primitive operations.

Changing data and formulas is done using the same primitive edit operations to
manipulate the document structure. A user-interface may provide a specialized editor, but still
trigger the pritive edits behind the scenes. However, interacting with elements in the document,
such as a entering text in a textbox, also generates a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

As we will see, past edits that demonstrate an operation done to the
document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
such recorded edits is implemented using the merging operation, which means that recorded operations
continue working even if the document structure later changes. Moreover, structural changes to the
document can be done independently and then merged (\S\ref{sec:impl-collab}). Evaluation of formulas
also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
edits done later by the user, the evaluated edits are removed, implementing an incremental
recomputation (\S\ref{sec:impl-incremental}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations.
\item We explain how a wide range of end-user programming experiences can be readily implemented using
  the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
\item We document important design decisions and alternatives (\S\ref{sec:discuss}). The analysis
  shows that the desired functionality requires a careful choice among interconnected design options.
\item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
  and evaluate the extent to which the Denicek substrate enables such development (\S\ref{sec:eval}).
\end{itemize}

\noindent
To enable others build on top of Denicek, we share our compact and documented
source code at: \url{https://github.com/d3sprog/denicek}

@@

\section{Background}
\label{sec:background}

What is this about? end-user programming \cite{myers-2006-eup}
but also programming systems \cite{jakubovic-2023-techdims},
Live \cite{rein-2019-live}, Interactive etc. - perhaps for a lack of better term use end-user
(also notebooks)

explain computational substrate
Substrate defined by \cite{jakubovic-2022-ladder}

citizen programmers? Maybe do not talk about ``programs'' because this is more documents with
formulas - not that fancy programs. Computational documents?

programming by demonstration
\cite{leiva-2021-rapido,cypher-1993-pbd}
\cite{chen-2023-miwa}

local first
\cite{kleppmann-2019-local,klokmose-2024-mywebstrates}

provenance/debugging
\cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
\cite{ricciotti-2017-imperative,perera-2012-functional}
\cite{perera-2022-linked}

end-user debugging
\cite{kissinger-2006-debugging}

live/incremental recomputation
\cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.

visualizing results of execution
projection boxes \cite{lerner-2020-boxes}

livelits (provide editors)
\cite{omar-2021-livelits}

edit calculus (based on tree navigation though)
\cite{omar-2017-hazelnut}

structure editing see \cite{beckman-2023-sandblocks}

Schema evolution
% our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2

Also on substrate see
% https://x.com/jonathoda/status/1185888711210389504

% also cool: https://arxiv.org/pdf/2303.06777

Webnicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
document visible to the user, although parts can be collapsible or hidden using CSS.

Merging - c.f. mywebstrates - we need more than Automerge (cite)
which has a bunch of pre-built structures

\cite{gobert-2023-lorgnette}

The Denicek substrate owes much to systems discussed in \S\ref{sec:background},
especially Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}.

@@

\begin{figure*}[t]
\vspace{-0.5em}
\includegraphics[width=0.95\textwidth,clip,trim=1cm 1cm 1cm 0.5cm]{fig/walkthrough.pdf}
\vspace{-1em}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

\newpage

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We follow two
formative examples (\S\ref{app:examples}) related to conference planning.

\subsubsection*{\circled{A} Adding a Speaker}
The user starts with an empty document, which is represented as a record. They add a field for
each heading and a list {\small\Verb_<ul>_}, represented by a field named {\small\Verb_speakers_}. They use the command
toolbox to add the first speaker.

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, the user creates a textbox and a button. They enter new speaker
details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
from the textbox into the new element using source view.

\subsubsection*{\circled{C} Saving the Interaction} After adding the speaker from the textbox,
the user opens history view, selects edits that added the speaker and saves those as the
{\small\Verb_add-speaker_} interaction. They attach this as a handler for the {\small\Verb_click_} event of the button.

\keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
demonstration (\S\ref{sec:impl-pbd}) by letting the user to save and replay past interactions,
either directly or by attaching them to a UI event.}

\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
version of the document and turns the list into a table. This is done by invoking a series of
commands that change tags, wrap elements, copy and transform values.

\subsubsection*{\circled{E} Merging Edits} The two document versions can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically added in the new format.

\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetic, but the order does not matter in the above scenario. The same merging
operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}

\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds budget calculation to
the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
are other nodes or references to other nodes or formulas (highlighted on mouse hover).

\keyideabox{\faLightbulbO}{Incremental Recomputation}{Evaluating a formula yields additional
edits that augment the document with the result. Those edits are kept
at the top of the document history and are removed in case of a conflict (\S\ref{sec:impl-incremental}),
providing incremental recomputation.}

@@

@@

>>>>

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

We contribute Denicek, a computational substrate that simplifies the implementation of the
above programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. It provides three
operations on edit histories: edit application, merging of histories and conflict resolution.
By composing these operations, we can  easily implement a range of programming experiences.

We present the architecture of Denicek, discuss key design considerations and elaborate
the implementation of a variety of programming experiences. To evaluate the proposed
substrate, we use Denicek to develop an innovative interactive data science notebook system.
The case study shows that the Denicek computational substrate provides a suitable basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

@@

A computational substrate defines the structures with which programs are constructed, how
the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

In this paper, we describe a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup}. Those include:

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and automatically merge their changes, preferrably without a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Stage Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for developing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

\subsubsection*{Substrate}
The Denicek substrate brings together two central design ideas. First, it represents programs as
document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
static content. Second, Denicek does not store the document tree itself, but instead, maintains
a sequence of edit operations through which the tree was constructed and transformed.

The substrate then provides three primitive operations for working with sequences of edits.
First, it can apply a series of edits to reconstruct the document. Second, it can merge two
diverging edit histories. Finally, it can detect conflicts when merging histories and, for
example, remove conflicting edits from one branch.

The key insight that we present in this paper is that many compelling programming experiences can
be implemented by leveraging the uniform document representation alongside with a suitable
composition of the three primitive operations.

Changing data and formulas is done using the same primitive edit operations that
manipulate the document structure. A user-interface may provide a specialized editor, but still
trigger the primitive edits behind the scenes. However, interacting with elements in the document,
such as a entering text in a textbox can also generate a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

As we will see, past edits that demonstrate an operation done to the
document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
such recorded edits is implemented using the merging operation, which means that recorded operations
continue working even if the document structure later evolves. Moreover, structural changes to the
document can be merged with concurrent data edits  (\S\ref{sec:impl-collab}). Evaluation of formulas
also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
edits done later by the user, the evaluated edits are removed, implementing an incremental
recomputation (\S\ref{sec:impl-incremental}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented using the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek (\S\ref{sec:case})
  and assess its impementation complexity~(\S\ref{sec:eval}).
\end{itemize}

\noindent
To enable others build on top of Denicek, we share our compact and documented
source code at: \url{https://github.com/removed/for/review} %d3sprog/denicek

@@

\section{Background}
\label{sec:background}
Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness, interactivity
and richness \cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
We focus on systems that allow gradual progression from a user to a developer, as envisioned
in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
part of this spectrum, also including spreadsheet systems and notebooks for data science.
We first review related systems and their technical design (\S\ref{sec:background-sys})
before discussing programming experiences (\S\ref{sec:background-exp}).

@@

\subsection{Programming Systems and Substrates}
\label{sec:background-sys}

\subsubsection*{Programming Systems and Substrates}
A number of systems illustrate the qualities Denicek aims to support and have a related internal
structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
use structured document-based program representation and provide some of the desired programming
experiences on top of this representation. Many of those design ideas can be traced back to
Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
sees is all there is) that we follow in Webnicek.

To indicate that Denicek is intended as an underlying infrastructure on top of which programming
systems can be built, we use the term \emph{computational substrate}, which also dates back to
Boxer~\cite{disessa-1995-epistemology} and is closely related to the notion of dynamic media
introduced by Kay and Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
is a substrate based on synchronization of documents (but without edit histories) that has been
use as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.

\subsubsection*{Edit Histories and Merging}
Manipulating programs through semantically meaningful edits is a technique used by structure
editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
edits have also been used as the basis of live programming environments \cite{storm-2013-deltas}.

Merging of edits is most frequently done in version control systems such as git. Systems based
on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

More generally, merging of edits can be based on the operational transform (OT) approach
\cite{davis-2002-otsgml}, where edit conflicts are reconciled,
or conflict-free representation (CRDTs) \cite{shapiro-2011-crdts,kleppmann-2019-local,litt-2022-peritext}.
In both approaches, supporting complex edits on tree structures
remains a challenge \cite{jungnickel-2016-otjson,da-2024-jsoncrdt}. Mergeable replicated data
types (MRDTs) \cite{kaki-2019-mrdts} can merge updates provided there is a suitable relational
representation of the data. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} use OT.

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

\subsection{Programming Experiences}
\label{sec:background-exp}

We review a range of compelling programming experiences explored in recent research,
focusing on those supported by Denicek.

\subsubsection*{Local-First Collaboration}
Since the early collaborative programming environments such as Collabode \cite{goldman-2011-collaborative},
real-time collaboration has became widely used, if not always without challenges \cite{tan-2024-vslive}.
Merging concurrent edits is one such challenge.
In addition to CRDT-based approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
collaboration have been solved through fine-grained locking \cite{wang-2024-nbconflicts}.

\subsubsection*{Programming by Demonstration}
Earliest PbD systems used the paradigm for tasks ranging from graphics
and user interfaces to general-purpose programming \cite{smith-1975-pygmalion,cypher-1993-pbd}.
Wrangler \cite{kandel-2011-wrangler} showed the effectivity of PbD for data cleaning,
whereas more recent uses range from augmented reality prototyping \cite{leiva-2021-rapido}
and web automation \cite{chen-2023-miwa} to end-user software customizations
\cite{litt-2020-customization,litt-2020-wildcard}. Nevertheless, expressing conditions in PbD remains an
active research topic~\cite{pu-2022-semanticon,radensky-2018-conditionals}. A more general class of demonstrational
interfaces \cite{myers-2000-demonstrational} also includes programming by example, used
for example to infer data transformations in spreadsheets \cite{gulwani-2012-spreadsheets}.
Demonstrational systems can be used to directly perform actions, but also to generate code
that captures the interaction, as done e.g. in WREX \cite{drossos-2020-wrex}.

\subsubsection*{Incremental Recomputation}
Interactive programming systems with live previews \cite{petricek-2020-live,mcdirmid-2013-usable}
have been attempting to update the previews without full recomputation at least since the pioneering
work on the Cornell Program Synthesizer \cite{teitelbaum-1981-cps}. Outside live programming,
more work has been focused on updating computation when data change, although this can also
be source code of a program passed to an adaptive interpreter \cite{acar-2006-adaptive}.
Incremental recomputation is also a concern in notebook systems where cells can be run out of
order \cite{singer-2020-jollity}. This can be addressed by maintaining a dependency graph
\cite{petricek-2018-wrattler,koop-2017-dataflow} that also allows incremental recomputation
on code change.

\subsubsection*{Schema Change Control}
Schema change control is concerned with adapting data and code to reflect changes in schema.
The problem is well-studied in the context of databases \cite{brahmia-2024-evolution},
although only few systems also automatically adapt database queries \cite{wang-2019-schema}.
However, the problem is starting to be recognized in programming systems research
\cite{litt-2020-cambria,edwards-2025-schema} as well as live programming \cite{barenz-2020-live}
where state needs to be preserved during program editing.

\subsubsection*{End-User Debugging}
Non-programmers also need to be able understand and debug their programs \cite{kissinger-2006-debugging}.
Systems such as Whyline and Probe Log \cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
record information about program execution to let users analyse why they see a particular result,
whereas displaying intermediate steps can aid understanding of data science pipelines \cite{shrestha-2021-unravel}.
More generally, such functionality can leverage provenance tracking \cite{cheney-2009-provenance}
and program slicing \cite{ricciotti-2017-imperative,perera-2012-functional}. The same infrastructure
can also be used to build linked visualizations \cite{perera-2022-linked}.

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
prototypes at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an additional
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
offer an appealing alternative design.

\subsubsection*{Other Programming Experiences}
Several experiences not directly addressed in this paper are worth further investigation.
Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
The problem of making live and rich editors compositional is addressed by
Engraft \cite{horowitz-2023-engraft}. Programming environments may also integrate
live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate. Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. In this section, we follow two
formative examples (\S\ref{app:examples}) related to conference planning.

\subsubsection*{\circled{A} Adding a Speaker}
The first user starts with an empty document, which is represented as a record. They add a field for
each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. They
use the command toolbox to add the first speaker.

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, the user creates a textbox and a button. They enter new speaker
details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
from the textbox into the new element using a source view.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
the user opens history view, selects edits that added the speaker and saves those as the
{\small\Verb_add-speaker_} interaction. They attach this interaction as a event handler for the
{\small\Verb_click_} event of the button.

\keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
demonstration (\S\ref{sec:impl-pbd}) by enabling the user to save and replay past interactions.
Those can be replayed directly, or used as event handlers in order to construct interactive user
interfaces (\S\ref{sec:impl-interaction}).}

\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
version of the document (A) and turns the list into a table. This is done by invoking a series of
commands that change tags, wrap elements, copy and transform values.

\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically added in the new format.

\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetric, but the order does not matter in the above scenario. The same merging
operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}

\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
are other nodes or references to other nodes or formulas (highlighted on mouse hover).

\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes and
evaluating a formula yields additional edits that augment the document with the result
(\S\ref{sec:impl-eval}). Those evaluated edits are kept at the top of the document history and are
removed in case of a conflict (\S\ref{sec:impl-incremental}), resulting in the incremental
recomputation behavior.}

@@

\section{The Denicek Substrate}
\label{sec:system}
Denicek represents programs as sequences of edits that construct and transform a computational
document. In this section, we describe the structure of documents and edits, as well as the
operations that form the backbone of the system and are used to implement the
end-user programming experiences as discussed in \S\ref{sec:impl}.

\subsection{Selectors, Documents and Edits}
\label{sec:system-defs}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
Records and lists are labeled with a \emph{tag} as in HTML. Record fields have unique names.
Lists are expected to contain elements of the same structure, identified by a unique index (serving
as an ID). Primitive nodes can be strings, numbers or references to another location in the
document tree. References can be relative or absolute. Edit operations only use absolute
references (to denote a target node), but relative references can appear in the document (e.g., to
refer to another node within a list element).

A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\ident{Any} selector makes it possible to refer to all elements of a list, but there is no
way to refer to all fields of a record. Note that Denicek does not use implicit numerical
indices for lists. The index of a new list item has to be supplied explicitly. The reasoning behind
this design choice is discussed in \S\ref{sec:discuss}.

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require a \emph{target} to which they are applied. Targets are absolute references not containing the
\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
and edits that add a new item take the index or field name of a previous node.

The edits allow an arbitrary transformation of a document via a series of steps whose effect can be
tracked by the substrate. As we saw, Denicek updates references when document structure changes.
Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
(above) and edits that affect references (below).

Renaming a field or wrapping a node updates any references to within the target location
(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
rejects the edit. A copy edit of a node to which there is a reference is also rejected, because it
is ambiguous whether references referring to the original location should be left unchanged,
or modified to point to the target of the copying (a reference cannot refer to two unrelated locations).

% \note{why not have distinct Copy and Move operations to resolve this?}
% TP: Good point - this is a nice motivation for Move! I should probably add that :-)

@@

\begin{figure}
\newcommand{\ektablecol}[5]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 \\[-0.2em]
&\multicolumn{2}{l|}{\sffamily\footnotesize #5}\\[0.3em]
}
\begin{tabular}{|cp{16em}r|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target \\[0.2em]
\hline
&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, after, node}{Record}
  {Add \textit{node} as a \textit{field} to the specified record \textit{after} a given field.}
\ektablecol{\faAt}{Append}{index, after, node}{List}
  {Append \textit{node} to the end of the specified list \textit{after} a given field.}
\ektablecol{\faSort}{Reorder}{permutation}{List}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faCode}{UpdateTag}{tag}{List or Record}
  {Change the tag of a specified list or record from to a new \textit{tag}.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}
  {Apply primitive \textit{transform} to the specified primitive.}
\hline
&&\\[-1em]
\ektablecol{\faFont}{RenameField}{old field, new field}{Record}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag, index}{Any}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.75em}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\vspace{-1em}
\end{figure}

\subsubsection*{Automatic Reference Update}
As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
statically enforced (the substrate can be seen as dynamically typed). As a consequence, edits
that transform the structure can be used as structural edits (affecting the structure)
or as value edits (affecting the value of specific nodes). The latter can be used, for example,
when constructing an additional list item. An item can be first constructed and then populated
with values. This temporarily violates the invariant that collections are homogeneous
(an issue we might address by using transactions as discussed in \S\ref{sec:discuss}).
Another use of value edits is when evaluating a formula, which involves wrapping the formula
node (\S\ref{sec:impl-eval}), but the operation should not affect references to the formula
itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
have a \emph{reference behavior} option that can be set to disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector,
which is necessarilly a value edit affecting a specific node.

% \note{I think it would be desireable to avoid all these complications. References
% should be automatically maintained so they ``just work''.}
% \note{I believe that evaluation should not change the type/shape of the document. The space
% for evaluation results should be be preallocated in some sense, either as extra structure in
% a formula, or as metadata attached to the tree. That said, I've tried various alternative
% approaches to this in Subtext without ever settling on a preferred solution.}
%
% TP: I think this is true - but that would make the system more "statically typed"
% and Denicek is dynamically typed - my only reason for this is to make the editing experience
% more "document-like" - but it is perhaps not necessary.

An important principle of the Denicek design is that the effect an edit has on references inside
the document does not depend on the current value of the document, only its type. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
% (The effect \ident{Copy} has on the document structure depends on the existing document
% structure, but not on its value).
This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
can be provided as an additional mechanism on top of the underlying Denicek substrate.

\subsection{Primitive Operations}
\label{sec:system-ops}
Denicek provides three primitive operations. A sequence of edits can be applied to a
document, two sequences of edits can be merged and they can be checked for conflicts. Denicek
identifies edit histories by a (git-like) hash, computed from the current edit and the hash of the
preceding edit. The hash is used to identify a common shared prefix of the
history during conflict resolution and merging.

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transforms it according to the edit. If
the edit affects references in the document (Fig~\ref{fig:edits}, below), Denicek updates
the relevant references according to the rules shown in Fig.~\ref{fig:updates}, provided
that the reference behavior of the edit is not set to disable this.
%(this is required when %the target reference contains the \ident{Index} selector).
